# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data

> Show any code that is needed to
> 
> 1. Load the data (i.e. `read.csv()`)
> 
> 2. Process/transform the data (if necessary) into a format suitable for your 
> analysis

The data is first unzipped and stored in the *data* directory. The resulting 
csv file is then read into a data frame using the `read.csv()` function.
```{r}
# Create a directory to store data, if one doesn't already exist
if(!file.exists("data/")){
    dir.create("data/")
}

# Unzip the data file
unzip(zipfile = "activity.zip", exdir = "data/")

# Load the data and create data frame 'activityDat'
activityDat <- read.csv("data/activity.csv")

# Remove missing values for the first part of the assignment
activityDatNonNA <- activityDat[!is.na(activityDat$steps), ]
```

We can use the `ddply()` function from the `plyr` library to create a new 
data frame with the number of steps listed by day.
```{r}
library(plyr)
dailyActivity <- ddply(activityDatNonNA, ~ date, summarize, 
                       steps = sum(steps))
```

Finally, to look at the daily activity _pattern_ below, we'll use `ddply()` 
again to create a data frame with the number of steps, averaged across all 
days, listed by interval.
```{r}
intervalActivity <- ddply(activityDatNonNA, ~ interval, summarize,
                          averageSteps = mean(steps))
```


## What is mean total number of steps taken per day?

> For this part of the assignment, you can ignore the missing values in the 
> dataset.
>
> 1. Make a histogram of the total number of steps taken each day
>
> 2. Calculate and report the **mean** and **median** total number of steps
> taken per day

The graphing package **ggplot2** can be used to make a histogram of the total 
number of steps taken each day.
```{r dailyStepsHistogram, fig.height = 4, fig.align = 'center'}
library(ggplot2)
ggplot(dailyActivity, aes(x = steps)) + 
    geom_histogram(aes(y = ..count..),
                   binwidth = 1000, colour = "black", fill = "#E69F00")
```

We can then calculate and report the mean and median total number of steps 
taken per day using the `mean()` and `median()` functions, respectively.
```{r}
dailyMean <- mean(dailyActivity$steps)
dailyMed <- median(dailyActivity$steps)
```
The mean total number of steps taken per day is 
**`r format(dailyMean, nsmall = 2)`**.  
The median number of steps per day is **`r dailyMed`**.


## What is the average daily activity pattern?

> 1. Make a time series plot (i.e. `type = "l"`) of the 5-minute interval 
> (x-axis) and the average number of steps taken, averaged across all days 
> (y-axis)
>
> 2. Which 5-minute interval, on average across all the days in the dataset, 
> contains the maximum number of steps?  

We'll use `ggplot()` again to make a line plot (which is equivalent to the 
`type = "l"` line plot)
```{r dailyPatternPlot, fig.height = 4, fig.align = 'center'}
ggplot(intervalActivity, aes(x = interval, y = averageSteps)) +
    geom_line(colour = "#0072B2")
```

The `which.max` function can be used to identify the interval with the maximum
number of steps (averaged across all days).
```{r}
intervalMax <- intervalActivity$interval[
    which.max(intervalActivity$averageSteps)]
```
The interval with the maximum number of steps is **`r intervalMax`**.


## Imputing missing values

> Note that there are a number of days/intervals where there are missing values
> (coded as `NA`). The presence of missing days may introduce bias into some 
> calculations or summaries of the data.
> 
> 1. Calculate and report the total number of missing values in the dataset 
> (i.e. the total number of rows with `NA`s)
>
> 2. Devise a strategy for filling in all of the missing values in the dataset.
> The strategy does not need to be sophisticated. For example, you could use 
> the mean/median for that day, or the mean for that 5-minute interval, etc.
>
> 3. Create a new dataset that is equal to the original dataset but with the 
> missing data filled in.
> 
> 4. Make a histogram of the total number of steps taken each day and 
> Calculate and report the **mean** and **median** total number of steps taken
> per day. Do these values differ from the estimates from the first part of the
> assignment? What is the impact of imputing missing data on the estimates of
> the total daily number of steps?

We can calculate the number of missing values using `is.na()`.
```{r}
numNA <- sum(is.na(activityDat$steps))
```

We'll use the mean number of steps for each interval (averaged across all days)
to fill in missing values for those intervals. The data frame 
'activityDatImputed' represents a new data set with the missing data filled in.
```{r}
# First identify NA entries corresponding to missing data
naEntries <- row.names(activityDat)[is.na(activityDat$steps)]
activityDatImputed <- activityDat

# For each missing entry, replace the NA with the average number of steps
# across all days for that entry's interval
for(entry in naEntries){
    avgIntervalActivity <- intervalActivity$averageSteps[
        intervalActivity$interval %in% activityDat[entry, "interval"]]
    activityDatImputed[entry, "steps"] <- avgIntervalActivity
}
```

Following the same approach from above, we'll use `ddply()` to create a new 
data frame with the number of steps listed by day.
```{r}
dailyActivityImputed <- ddply(activityDatImputed, ~ date, summarize, 
                       steps = sum(steps))
```

We can make another histogram with `ggplot()` to look at the distribution of 
steps taken each day.
```{r dailyStepsHistogram_imputed, fig.height = 4, fig.align = 'center'}
ggplot(dailyActivityImputed, aes(x = steps)) + 
    geom_histogram(aes(y = ..count..),
                   binwidth = 1000, colour = "black", fill = "#E69F00")
```

Finally, we'll calculate the mean and median as before.
```{r}
dailyMeanImputed <- mean(dailyActivityImputed$steps)
dailyMedImputed <- median(dailyActivityImputed$steps)
```
The new mean total number of steps taken per day is 
**`r format(dailyMeanImputed, nsmall = 2)`**, and the median number of steps 
per day is **`r dailyMed`**. Compared to the previous mean and median, the mean 
number of steps per day with the imputed data is unchanged; however, the median 
is now equal to the mean.


## Are there differences in activity patterns between weekdays and weekends?

> For this part the `weekdays()` function may be of some help here. Use the
> dataset with the filled-in missing values for this part.
>
> 1. Create a new factor variable in the dataset with two levels – “weekday”
> and “weekend” indicating whether a given date is a weekday or weekend day.
>
> 2. Make a panel plot containing a time series plot (i.e. `type = "l"`) of the
> 5-minute interval (x-axis) and the average number of steps taken, averaged
> across all weekday days or weekend days (y-axis). 

We'll create a new factor variable in the dataset with two levels – “weekday” 
and “weekend” (indicating whether a given date is a weekday or weekend day)
using multiple steps with `weekdays()`, `%in%`, and `as.factor()`.
```{r}
# Create a logical vector indicating whether each day represents a weekday or 
# weekend using the 'weekends' function
weekendDays <- c("Saturday", "Sunday")
isWeekday <- !(weekdays(as.Date(activityDatImputed$date)) %in% weekendDays)

# Create a new variable in 'activityDatImputed' with entries of "weekday" or 
# "weekend" corresponding to values of TRUE or FALSE in 'isWeekend', 
# respectively
activityDatImputed$day[isWeekday] <- "weekday"
activityDatImputed$day[!isWeekday] <- "weekend"

# Convert the 'day' variable to a two-level factor using 'as.factor'
activityDatImputed$day <- as.factor(activityDatImputed$day)
```

To again look at the daily activity _pattern_, we'll use `ddply()` to create a 
data frame with the number of steps, averaged across all days, listed by 
interval; note: this time there will be one entry for the weekday average and
one for the weekend average.
```{r}
intervalActivityImputed <- ddply(activityDatImputed, .(interval, day), 
                                 summarize, averageSteps = mean(steps))
```

Finally, to compare activity patterns between weekdays and weekend days, we'll
use `ggplot()` to create two line plots, stacked vertically.
```{r dailyPatternPlot_comparison, fig.height = 6, fig.align = 'center'}
ggplot(intervalActivityImputed, aes(x = interval, y = averageSteps)) +
    geom_line(colour = "#0072B2") +
    facet_wrap(~ day, nrow = 2)
```

Weekend days appear to have more steps per interval on average, but with fewer
steps during morning invervals.
