---
title: "Motor Trend mtcars Analysis"
author: "James Eddy"
date: "August 24, 2014"
output: pdf_document
---

### Executive Summary

The objective of this study was to investigate the following questions:
> “Is an automatic or manual transmission better for MPG”
> "Quantify the MPG difference between automatic and manual transmissions"

After exploratory calculations and plotting to inspect apparent relationships
between transmission and MPG (and between each and other variables in the data),
several linear models were fit and examined. While transmission type appears
to have a significant effect on MPG when viewed in isolation, further analysis
revealed that this result is likely due to confounding from other variables.
In particular, transmission type is negatively correlated with car weight,
which has a significant negative association with MPG, thus creating the
apparent (but non-significant) positive association between transmission and
gas mileage.


### Exploring the Data

The data for this study can be found in the `mtcars` dataset within **R**. 
Exploratory analysis was performed using functions in the `dplyr` and `ggplot2`
libraries. After loading the required data and libraries, the `cor()` function
was used to examine the correlation between variables in the data set. This
simple measure suggested that transmission type is only moderately correlated
with MPG (0.60), while other values such as number of cylinders, displacement, 
and weight might have more of an effect on gas mileage.

```{r load, echo=FALSE}
library(dplyr)
library(ggplot2)
library(reshape2)

data(mtcars)
mtcarsNum <- mtcars
cor(mtcarsNum)[c("mpg", "am"), ]
```

Before generating some exploratory plots, the `mutate()` function was used to 
append an additional column `transmission` (based on the `am` column) to the 
`mtcars` data frame, making inspection of automatic vs. manual transmission 
easier. Generating boxplots of MPG values for automatic and manual
transmission cars provided a simple way to compare distributions (Figure 1).
The difference in median MPG shown in the plots suggested that manual
transmission cars are better for gas mileage than automatic transmission.
While there was a pronounced difference in MPG between manual and automatic 
transmission cars, there may be other factors correlated with transmission
that are more responsible for the difference in MPG. Several variables,
including cylinder, displacement, rear axle ratio, weight, and number of gears
all appear to have either positive or negative associations with transmission
type (Figure 2). These variables represent potentially "confounding" factors,
which need to be controlled for when modeling the influence of transmission 
type on MPG.

``` {r mutate, echo=FALSE}
mtcars <- mtcars %>%
    mutate(transmission = factor(am, labels = c("automatic", "manual"))) %>%
    select(-am)
```

### Modeling

A simple model was first fit to directly measure the apparent influence of
transmission on MPG. The first non-intercept coefficient of the model,
7.245, suggested an increase of 7.245 miles per gallon when switching from
automatic to manual transmission (with p-value < 0.0001). However, the low R-
squared value (0.3598) of the model indicated that transmission alone is not a 
very good predictor of the variation in MPG.

```{r model1, echo=TRUE}
fm1 <- formula(mpg ~ transmission)
model1 <- lm(fm1, data = mtcars)
summary(model1)
```

To examine whether additional attributes may be confounding the observed
influence of transmission type on MPG, a second model was fit with three
additional independent variables: rear axle ratio (`drat`), weight (`wt`), and
number of gears (`gear`). Each of these variables was previously shown in the
exploratory analysis to be highly correlated with transmission type. Notably,
the resulting fit indicates that transmission is in fact not a strong predictor
of MPG, with only 0.1689 (p-value = 0.94) increase in MPG for manual vs. 
automatic. Instead, weight seems to have a much stronger relationship to gas
mileage, with a decrease of 4.89 miles-per-gallon predicted for every increase
of 1000 lbs in car weight.

Even with the four indepentent variables, the second model still only led to a
moderately good R-squared value of 0.7687. Furthermore, a diagnostic plot of 
residuals showed values non-evenly distributed around zero (Figure 3), 
indicating that the model is probably not very good. To examine whether adding 
additional variables could improve predictability, a third model was fit based 
on all variables in the data set. While R-squared was improved (0.869), the
plot of residuals was still somewhat uneven. It is possible that too many
variables are included in this case, which might hurt the model; alternatively,
there may be interaction among some variables, which could be modeled more
explicitly to improve performance. As the objective of this assignment was to
explore the effect of transmission type on MPG, further optimization of a model
for MPG prediction was not pursued here.

```{r model2, echo=FALSE}
fm2 <- formula(mpg ~ transmission + drat + wt + gear)
model2 <- lm(fm2, data = mtcars)
```

```{r modelAll, echo=FALSE}
fmAll <- formula(mpg ~ .)
modelAll <- lm(fmAll, data = mtcars)
```


### Appendix

```{r plot1, echo=FALSE}
ggplot(mtcars, aes(y = mpg, x = transmission)) +
    geom_boxplot(aes(fill = transmission)) +
    geom_jitter(aes(size = mpg), alpha = 0.5) +
    labs(title = "Figure 1")
```


```{r plot2, echo=FALSE}
mtcars_long <- melt(mtcars, id.vars = c("transmission", "mpg"))

ggplot(mtcars_long, aes(y = value, x = transmission)) +
    geom_boxplot(aes(fill = transmission)) +
    geom_jitter(aes(size = mpg), alpha = 0.5) +
    scale_x_discrete(breaks = NULL) +
    facet_wrap(~ variable, scales = "free_y", nrow = 2) +
    labs(title = "Figure 2")
```

```{r resid1, echo=FALSE}
residPred <- data.frame(residuals = residuals(model2), yhat = predict(model2))

ggplot(residPred, aes(x = yhat, y = residuals)) +
    geom_point() +
    xlab("Fitted values") +
    ylab("Residuals") +
    geom_hline(yintercept = 0) +
    geom_smooth(method = "loess") +
    labs(title = "Figure 3")
```

```{r resid2, echo=FALSE}
residPred <- data.frame(residuals = residuals(modelAll), 
                        yhat = predict(modelAll))

ggplot(residPred, aes(x = yhat, y = residuals)) +
    geom_point() +
    xlab("Fitted values") +
    ylab("Residuals") +
    geom_hline(yintercept = 0) +
    geom_smooth(method = "loess") +
    labs(title = "Figure 3")
```

